---
title: "Master's Thesis Code - Data Pre-processing"
author: "Ana Pérez Barrera"
date: "`r Sys.Date()`"
output: html_document
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

```{r setup, include=FALSE}
knitr::opts_knit$set(echo=TRUE, warning=FALSE,message=FALSE)
```

```{r, include=FALSE}
library(vroom)
library(car)
library(caret)
library(cowplot)
library(corrplot)
library(dplyr)
library(ggplot2)
library(ggcorrplot)
library(ggpubr)
library(gbm)
library(glmnet)
library(mice)
library(patchwork)
library(pdp)
library(randomForest)
library(readr)
library(reshape)
library(scales)
library(stringr)
library(sf)
library(tidyverse)
library(tidyr)
```

## 1. LABOR FORCE SURVEY DATA PRE-PROCESSING

```{r, eval=FALSE}
# Function to process CSV files for a given country
process_country <- function(country_code, folder_path) {
  # Set the path to the country's folder containing the CSV files
  country_folder <- paste0(folder_path, "/", country_code)
  
  # List all CSV files in the country's folder
  csv_files <- list.files(country_folder, pattern = "\\.csv$", full.names = TRUE)
  
  # Initialize an empty list to store data frames
  data_list <- list()
  
  # Loop through each CSV file
  for (csv_file in csv_files) {
    # Read the CSV file and select the desired columns
    if (file.exists(csv_file)) {
      year <- substr(csv_file, nchar(csv_file) - 9, nchar(csv_file) - 6)
      
      if (!is.na(year) && as.numeric(year) < 2011) {
        df <- vroom(csv_file, col_select = c("COUNTRY", "YEAR", "ISCO88_3D", 
                                             "SEX", "EMPSTAT", "HOMEWORK", "FTPT", 
                                              "SUPVISOR", "HWUSUAL", "EXTRAHRS", "SHIFTWK",
                                             "EVENWK", "NIGHTWK","SATWK", "SUNWK"))
      } else {
        df <- vroom(csv_file, col_select = c("COUNTRY", "YEAR", "ISCO08_3D", 
                                             "SEX", "EMPSTAT", "HOMEWORK", "FTPT", 
                                             "SUPVISOR", "HWUSUAL","EXTRAHRS","SHIFTWK",
                                             "EVENWK", "NIGHTWK","SATWK", "SUNWK"))
      }
      
      # Store the data frame in the list
      data_list[[csv_file]] <- df
    }
  }
  
  # Standardize column names across all data frames in data_list
  data_list <- lapply(data_list, function(df) {
    colnames(df) <- c("COUNTRY", "YEAR", "ISCO", 
                      "SEX", "EMPSTAT", "HOMEWORK", "FTPT", 
                      "SUPVISOR", "HWUSUAL","EXTRAHRS",
                      "SHIFTWK","EVENWK", "NIGHTWK", 
                      "SATWK", "SUNWK")
    return(df)
  })
  
  # Combine data frames from all CSV files into a single data frame
  country_final <- do.call(rbind, data_list)
  
  # Write the final dataset to a CSV file
  write.csv(country_final, paste0(country_code, ".csv"), row.names = FALSE)
}

# Specify the folder path
folder_path <- "C:/Users/LG/Desktop/TFM/LFS - microdata"

# List of country codes
country_codes <- c("BE-Belgium","DE-Germany","DK-Denmark","EL-Greece",
                   "ES-Spain","FI-Finland", "FR-France","IT-Italy",
                   "NL-Netherlands","NO-Norway","PT-Portugal","SE-Sweden"
                   "AT-Austria", "IE-Ireland", "LU-Luxembourg")

# Process each country
for (country_code in country_codes) {
  process_country(country_code, folder_path)
}

```


Fuction created to perform the ISCO transformation from 3 to 2 digit

```{r, eval=FALSE}
# Create a function to extract the first two digits of ISCO 3-digit codes
extract_ISCO <- function(isco3d) {
  substr(isco3d, 1, 2)
}
```

### Handling missing values in LFS

Iteration over each COUNTRY/dataset to clean and process the data: ISCO Transformation and transforming missing values. Save one final processed file for each country

```{r, eval=FALSE}
# List all CSV files in your working directory
csv_files <- list.files(pattern = "\\.csv$")

# Iterate over each CSV file
for (file in csv_files) {
  # Read the CSV file
  data <- read.csv(file)
  
  # ISCO transformation from 3 to 2 digit
  # Apply the fucntion to ISCO variable (2-digit extraction)
  data$isco <- extract_ISCO(data$ISCO)
  data <- data |> dplyr::select(-ISCO)
  # Convert ISCO 99 in NA
  data$isco[data$isco == 99] <- NA
  
  # Being in employment: EMPSTAT
  # Removing Out of labor force (EMPSTAT=9)
  data$EMPSTAT[data$EMPSTAT == 9] <- NA
  data <- data[!is.na(data$EMPSTAT), ]
  # Removing not employed (EMPSTAT=2)
  data <- subset(data, EMPSTAT != 2)
  
  # Independent variables: labour market participation
  data$HOMEWORK[data$HOMEWORK == 9] <- NA
  data$FTPT[data$FTPT == 9] <- NA
  data$SUPVISOR[data$SUPVISOR == 9] <- NA
  
  # Independent variables: working conditions (working hours and time arrangements)
  data <- data[, names(data) != "HWUSUAL"]
  data$EXTRAHRS[data$EXTRAHRS == 99] <- NA
  data$SHIFTWK[data$SHIFTWK == 9] <- NA
  data$EVENWK[data$EVENWK == 9] <- NA
  data$NIGHTWK[data$NIGHTWK == 9] <- NA
  data$SATWK[data$SATWK == 9] <- NA
  data$SUNWK[data$SUNWK == 9] <- NA

   # Create the new file name with "_final"
  final_file <- sub("\\.csv$", "_final.csv", file)
  
  # Save the modified data back to CSV file with the new name
  write.csv(data, file = final_file, row.names = FALSE)
  
}
```

##  2. JOINT RESEARCH CENTER DATA PRE-PROCESSING

```{r, eval=FALSE}
tasks <- read_delim("tasks eu isco2d nace2d EU15.csv", 
                     delim = "\t", escape_double = FALSE, 
                     trim_ws = TRUE)

# Save vector of unique occupations before removing them
occupation <- unique(tasks$isco)

isco_df <- data.frame(
  isco = as.numeric(gsub("\\..*", "", occupation)),
  job = gsub("^\\d+\\.\\s*", "", occupation)
)

# Save isco_df for later
# write.csv(isco_df, "isco_df.csv", row.names = F)
```

### NA imputation

```{r, eval=FALSE}
# Remove text from ISCO column
tasks$isco <- gsub("[^0-9]+", "", tasks$isco)
# Removing nace and pop variables
tasks<- tasks |>  dplyr::select (-nace,-pop)

#NA imputation with MICE method
set.seed(123)
m = 5 
mice_mod <- mice(tasks, m=m, method='rf')
isco2d <- complete(mice_mod, action=m)

sapply(isco2d, function(x) sum(is.na(x))*100/nrow(isco2d))

# Save isco2d for later
write.csv(isco2d, "isco2d.csv", row.names = F)
```

**Define Country Data**

```{r, eval=FALSE}
country_data <- list(
  AT_Austria="AT-Austria_final.csv",
  BE_Belgium = "BE-Belgium_final.csv",
  DE_Germany = "DE-Germany_final.csv",
  DK_Denmark = "DK-Denmark_final.csv",
  EL_Greece = "EL-Greece_final.csv",
  ES_Spain = "ES-Spain_final.csv",
  FI_Finland = "FI-Finland_final.csv",
  FR_France = "FR-France_final.csv",
  IE_Ireland ="IE-Ireland_final.csv",
  IT_Italy = "IT-Italy_final.csv",
  LU_Luxembourg="LU-Luxembourg_final.csv",
  NL_Netherlands = "NL-Netherlands_final.csv",
  NO_Norway = "NO-Norway_final.csv",
  PT_Portugal = "PT-Portugal_final.csv",
  SE_Sweden = "SE-Sweden_final.csv"
)
```

### Ratios by country-occupation-year

#### A. Female ratio

Create female ratio from 0 (fully masculine occupation) to 1 (fully feminine occupation).

```{r, eval=FALSE}
female_ratio <- function(data) {
data <- data %>%
    group_by(YEAR, isco) %>%
    summarise(
      male_count = sum(SEX == 1,  na.rm = TRUE),
      female_count = sum(SEX == 2,  na.rm = TRUE) ) %>%
    mutate(
      total = male_count + female_count,
      femratio = female_count / total ) %>%
    mutate(
      femratio = round(femratio, 4)) |> 
    select(YEAR, isco, femratio) |> 
   left_join(data, by = c("YEAR", "isco")) |> 
    dplyr::select(-SEX)

return(data)
} 
```

#### B. Independent variables: labor market participation

-   **HOMEWORK**: Working at home for the main job (EMPSTAT=1): 1 Person mainly works at home, 2 Person sometimes works at home, 3 Person never works at home.

Create an online-work ratio from 0 (fully in person) to 1 (fully digitalized/online work)

In this code, the total sum of persons working at home is calculated, considering that persons working both at home and in the office (both_count) have a weight of 0.5. Then, the work-at-home ratio is calculated by dividing the sum of the count of persons working exclusively at home and half of the count of persons working at home and in the office by the total.

```{r, eval=FALSE}
online_ratio <- function(data) {
  data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      online_count = sum(HOMEWORK == 1, na.rm = TRUE),
      both_count = sum(HOMEWORK == 2, na.rm = TRUE), 
      office_count =sum(HOMEWORK ==3, na.rm = TRUE)) |> 
    mutate(
      total = online_count + both_count * 0.5 + office_count,
      onratio = (online_count + both_count * 0.5) / total) |> 
    mutate(onratio = round(onratio, 4)) |>
      select(YEAR, isco, onratio) |> 
     left_join(data, by = c("YEAR", "isco")) |> 
     dplyr::select(-HOMEWORK)
  
  return(data)
} 
```

-   **FTPT**: Full- or part-time main job (self-defined) (EMPSTAT=1): 
- 1 Full-time job
- 2 Part-time job

Create full-time/part-time ratio from 0 (part-time) to 1 (full-time).

```{r, eval=FALSE}
ftpt_ratio <- function(data) {
  data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      ft_count = sum(FTPT == 1, na.rm = TRUE),
      pt_count = sum(FTPT == 2, na.rm = TRUE)) |> 
    mutate(
      total = ft_count + pt_count,
      ftratio = ft_count / total) |> 
    mutate(ftratio = round(ftratio, 4)) |> 
    select(YEAR, isco, ftratio) |> 
    left_join(data, by = c("YEAR", "isco")) |> 
    dplyr::select(-FTPT)
  
  return(data)
}
```

-   **SUPVISOR**: Supervisory responsibilities in main job
-   1 Yes
-   2 No

Create supervisor responsability ratio from 0 (no supervisory responsabilites) to 1 (supervisory responsabilities).

```{r, eval=FALSE}
supvisor_ratio <- function(data) {
  
  data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      supvisor_yes = sum(SUPVISOR == 1, na.rm = TRUE),
      supvisor_no = sum(SUPVISOR == 2, na.rm = TRUE)) |> 
    mutate(
      total = supvisor_yes + supvisor_no,
      supratio = supvisor_yes / total) |> 
    mutate(supratio = round(supratio, 4)) |> 
    select(YEAR, isco, supratio) |> 
    left_join(data, by = c("YEAR", "isco")) |> 
    dplyr::select(-SUPVISOR)
  
  return(data)
}
```

#### B. Independent variables: working conditions

-   **EXTRAHRS**: Overtime or extra hours worked in main job: 0 No overtime or extra hours in the main job, 0.5-95, Number of hours of overtime or extra hours in the main job (with one possible decimal for half hours). We create a ratio for:
-   0 - None Extra hours.
-   1 - Yes Extra hours

```{r, eval=FALSE}
extrahrs_ratio <- function(data) {
  
  data <- data |> 
    mutate( EXTRAHRS = case_when(
    EXTRAHRS == 0 ~ "0",
    EXTRAHRS < 20 ~ "1"
  ))
 
   data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      extra_no = sum(EXTRAHRS == 0,  na.rm = TRUE),
      extra_yes = sum(EXTRAHRS == 1,  na.rm = TRUE)) |> 
    mutate(
      total = extra_no + extra_yes,
      extraratio = extra_yes / total) |> 
    mutate(extraratio = round(extraratio, 4)) |> 
      select(YEAR, isco, extraratio)|> 
    left_join(data, by = c("YEAR", "isco")) |> 
    dplyr::select(-EXTRAHRS)
  
  return(data)
}
```

-   **SHIFTWK**: Shift work in main job 
- 1. Person usually does shift work. 
- 3. Person never does

```{r, eval=FALSE}
shiftwk_ratio <- function(data) {
 
   data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      shift_no = sum(SHIFTWK == 3,  na.rm = TRUE),
      shift_yes = sum(SHIFTWK == 1,  na.rm = TRUE)) |> 
    mutate(
      total = shift_no + shift_yes,
      shiftratio = shift_yes / total) |> 
    mutate(shiftratio = round(shiftratio, 4)) |> 
      select(YEAR, isco, shiftratio)|> 
    left_join(data, by = c("YEAR", "isco")) |> 
      dplyr::select(-SHIFTWK)
  
  return(data)
}
```

-   **EVENWK**: Evening work in main job

-   **NIGHTWK**: Night work in main job

-   **SATWK**: Saturday work in main job

-   **SUNWK**: Sunday work in main job

1- Person frequently / 2- Person sometimes / 3- Person never

```{r, eval=FALSE}
shifts_ratio <- function(data) {
  
  data <- data |> 
    group_by(YEAR, isco) |> 
    summarise(
      freq_count = sum(EVENWK == 1, na.rm = TRUE),
      both_count = sum(EVENWK == 2, na.rm = TRUE), 
      never_count =sum(EVENWK ==3, na.rm = TRUE)) |> 
    mutate(
      total = freq_count + both_count * 0.5 + never_count,
      evenratio = (freq_count + both_count * 0.5) / total) |> 
    mutate(evenratio = round(evenratio, 4)) |>
      select(YEAR, isco, evenratio) |> 
     left_join(data, by = c("YEAR", "isco"))
  
    data <- data %>%
      group_by(YEAR, isco) |> 
    summarise(
      freq_count = sum(NIGHTWK == 1, na.rm = TRUE),
      both_count = sum(NIGHTWK == 2, na.rm = TRUE), 
      never_count =sum(NIGHTWK ==3, na.rm = TRUE)) |> 
    mutate(
      total = freq_count + both_count * 0.5 + never_count,
      nightratio = (freq_count + both_count * 0.5) / total) |> 
    mutate(nightratio = round(nightratio, 4)) |>
      select(YEAR, isco, nightratio) |> 
     left_join(data, by = c("YEAR", "isco"))
   
    data <- data %>%
      group_by(YEAR, isco) |> 
      summarise(
      freq_count = sum(SATWK == 1, na.rm = TRUE),
      both_count = sum(SATWK == 2, na.rm = TRUE), 
      never_count =sum(SATWK ==3, na.rm = TRUE)) |> 
    mutate(
      total = freq_count + both_count * 0.5 + never_count,
      satratio = (freq_count + both_count * 0.5) / total) |> 
    mutate(satratio = round(satratio, 4)) |>
      select(YEAR, isco, satratio) |> 
     left_join(data, by = c("YEAR", "isco"))
    
    data <- data %>%
      group_by(YEAR, isco) |> 
      summarise(
      freq_count = sum(SUNWK == 1, na.rm = TRUE),
      both_count = sum(SUNWK == 2, na.rm = TRUE), 
      never_count =sum(SUNWK ==3, na.rm = TRUE)) |> 
    mutate(
      total = freq_count + both_count * 0.5 + never_count,
      sunratio = (freq_count + both_count * 0.5) / total) |> 
    mutate( sunratio = round(sunratio, 4)) |>
      select(YEAR, isco, sunratio) |> 
     left_join(data, by = c("YEAR", "isco")) |> 
      dplyr::select(-EVENWK, -NIGHTWK, -SATWK, -SUNWK)
  
  return(data)
} 
```

#### C. Proportion of employees by country, occupation and year.

This function is used to calculate the ISCOPROP variable, that represents the proportion of employees represented in each occupation classification, out of the total nº of employees, per year. Each row in prop_data represents the proportion of employees in a specific occupation for a specific year. 

```{r, eval=FALSE}
# Function to calculate proportions by occupation and year
calculate_iscoprop <- function(data) {
  prop_data <- data |> 
    group_by(YEAR,isco) |> 
    summarise(employed = sum(EMPSTAT == 1)) |> 
    ungroup() |> 
    group_by(YEAR) |> 
    mutate(total = sum(employed)) |> 
    mutate(iscoprop = (employed / total) * 100) |> 
    dplyr::select(-employed, -total) 
  
  return(prop_data)
}

# Add ISCOPROP in the main dataframe
add_iscoprop <- function(data) {
  data <- data |> left_join(prop_data,by = c("YEAR", "isco"))
  return(data)
}
```

Then, the following function calculates the variation in proportions

```{r, eval=FALSE}
total_employees <- function(data) {
data <- data |> 
  group_by(YEAR,isco) |> 
  summarise(totalemp = sum(EMPSTAT == 1)) |> 
  left_join(data, by = c("YEAR","isco"))

return(data)
}
```

**Let's apply all the functions**

```{r, eval=FALSE}
for (country in names(country_data)) {
  filename <- country_data[[country]]
  cntdata <- read_csv(filename)
  cntdata <- female_ratio(cntdata)
  cntdata <- online_ratio(cntdata)
  cntdata <- ftpt_ratio(cntdata)
  cntdata <- supvisor_ratio(cntdata)
  cntdata <- extrahrs_ratio(cntdata)
  cntdata <- shiftwk_ratio(cntdata)
  cntdata <- shifts_ratio(cntdata)
  prop_data <- calculate_iscoprop(cntdata)
  cntdata <- add_iscoprop(cntdata)
  cntdata <- total_employees(cntdata)
  
  assign(country, cntdata)
  
}
```

Let's use the `distinct` function to keep only the unique values in the dataset. **Note that** each observation represents a unique year-occupation combination.

```{r, eval=FALSE}
# Lista para almacenar los nombres de los conjuntos de datos procesados
processed_datasets <- list()

# Iterar sobre los objetos en el entorno
for (obj_name in ls()) {
  # Verificar si el objeto es un dataframe y tiene el prefijo "country_code_"
  if (is.data.frame(get(obj_name)) && grepl("^[A-Z]{2}_", obj_name)) {
    # Obtener el dataframe
    data <- get(obj_name)
    
    # Apply distinct function
      data <- data %>%
      distinct(.keep_all = TRUE) 
    
    # Guardar el dataframe procesado en la lista
    processed_datasets[[obj_name]] <- data
  }
}

# Asignar los conjuntos de datos procesados a nuevos objetos en el entorno
for (obj_name in names(processed_datasets)) {
  assign(obj_name, processed_datasets[[obj_name]])
}
```

Let's merge all country datasets (2006-2021) and get rid of unnecessary noise in the isco variable.

```{r}
merged_data <- bind_rows(processed_datasets)
# Remove non-categorized values in job-isco
merged_data$isco[merged_data$isco %in% c(10,20,30,40,50,60,70,80,90,0)] <- NA
merged_data <- merged_data[!is.na(merged_data$isco), ]
```

```{r, eval=FALSE}
write.csv(merged_data, file = "merged_data.csv", row.names  = FALSE)
```





